---
title: "Appendix"
description: "Connecticut Real Estate Sales Data"
author: "Joseph Farrell"
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


##Appendix 

__Below is the code up to line 132 in FINAL PAPER DOC__
```{r}
library(readr)
library(tidyverse)
library(dplyr)
library(mosaic)
library(ggplot2)
library(lubridate)
library(xts)
library(formattable)
library(forecast)
library(kableExtra)
library(plotly)
library(gapminder)
library(gganimate)
library(gifski)
library(png)

connecticut <- read_csv("/Users/nelsonfarrell/Downloads/Real_Estate_Sales_2001-2019_GL.csv")
county_town <- read_csv(("/Users/nelsonfarrell/Documents/501 Stats/Connecticut County:Towns.csv"))
county_town <- county_town %>%
  select("subregion", "town")

connecticut <- rename(connecticut, "town" = "Town")

connecticut <- 
  left_join(connecticut, county_town, by = "town")

connecticut <- select(connecticut, 
                      "List Year", 
                      "Date Recorded", 
                      "town", 
                      "Assessed Value", 
                      "Sale Amount",
                      "Residential Type",
                      "subregion")

connecticut <- connecticut %>%
  na.omit(`Residential Type`)

connecticut <- rename(connecticut, 
                      "list_year" = "List Year", 
                      "sale_date" = "Date Recorded", 
                      "assessed_price" = "Assessed Value", 
                      "sale_price" = "Sale Amount",
                      "property_type" = "Residential Type", 
                      "county" = "subregion")

connecticut$sale_date <- as_date(connecticut$sale_date, 
                                 format = "%m/%d/%Y")
connecticut <- connecticut %>%
  mutate(sale_year = year(sale_date))

```


__Examine Counts in relation to year__
```{r}
connecticut %>%
  group_by(sale_year) %>%
  count(sale_year) %>%
  mutate("Proportion of Total" = n/548175) %>%
  mutate(across(contains("Proportion"), round, 2)) %>%
  rename("Sale Year" = "sale_year", 
         "Properties Sold" = "n") %>%
  knitr::kable(align = c("l", "c", "r"),
               caption = "Number of Properties Sold Each Year ") %>%
  kable_material(bootstrap_options = c("striped", "hover"), full_width = T) %>%
  row_spec(1:6, background = "yellow") 
```

```{r}
connecticut %>%
  group_by(sale_year) %>%
  ggplot(aes(x = sale_year)) +
  geom_bar(fill = "steelblue")
```

The table and the bar graph display the scope of the data. While there are potenitally enough observations 
to keep the year 2006 (10,977 observations); it is still considerablly less than the years following. The minimum year
total after 2006 is 2011 (28,892 observations) which makes up approxiamtely 5% of the data. 2006 only makes up 2%.

As a result of the data prior to and including 2006 being very limited, I will remove all observations prior to 2007.

__Remove observations prior to 2006 and including 2006__
```{r}
connecticut <- connecticut %>%
  filter(sale_year > 2006)
```

## Distribution of Sale Price

__Get summary statistics of `assessed_price` and `sale_price`__
```{r}
summarize(connecticut,
          "Mean Sale Price" = mean(sale_price),
          "Median Sale Price" = median(sale_price),
          "Standard Deviation Sale Price" = sd(sale_price))
```

__View favstats of `sale_price`__
```{r}
options(scipen = 999) # remove scientific notation
favstats(~sale_price, data = connecticut)
```

__View summary statistics of different `residential_types`__
```{r}
connecticut %>%
  group_by(property_type) %>%
  summarize("Mean Sale Price" = mean(sale_price),
            "Median Sale Price" = median(sale_price),
            "Standard Deviation Sale Price" = sd(sale_price)) %>%
  rename("Proerty Type" = "property_type") %>%
  knitr::kable(caption = "Distribution of Sale Price") %>%
  kable_material(full_width = T)
  
  
```

__View distribution of `sale price`, check for outliers__
```{r}
connecticut %>%
  group_by(property_type) %>%
  ggplot(aes(x = property_type, 
             y = sale_price)) +
  geom_violin(notch = TRUE) +
  labs(title = "Distribution of Sale Price for Residential Types",
       x = "Property Type",
       y = "Sale Price") +
  theme_light()
```

* The violin plots reveal that the data is heavily skewed to the right.
* The summary statistics reveal very high standard deviations, 
  and medians and means that are not close in value.
* For these reasons I will effectively use trimmed mean as a measure of central 
  tendency but I will actually remove the highest and lowest 2.5% of 
  observations (sale_price). 

__Remove top and bottom 2.5% of "sale_price"__
```{r}
connecticut <- connecticut %>%
  filter(sale_price < quantile(connecticut$sale_price, .975) & 
           sale_price > quantile(connecticut$sale_price, .025)) 
537181-509278
```

 
 __View favstats of `sale_price` after being trimmed__  
```{r}
favstats(~sale_price, data = connecticut)
```

* Here we see that standard has been reduced from $608,700 to $216,198.4 indicating
  that the dispersion has been decreased.
* The mean reduced from $339,121.6 to $283,812. and is now closer to the median.
* The minimum has increased from $0 to $33,697, and the maximum has decreased from 
  $157,000,000 to $1,449,000 indicating the removal of at least some the most
  extreme outliers.
* Trimming the top and bottom 2.5% removed a total of 27,903 observations.
  
  __View distribution after trimming the top and bottom 2.5%__
```{r}
connecticut %>%
  group_by(property_type) %>%
  ggplot(aes(x = property_type, 
             y = sale_price)) +
  geom_violin(aes(fill = property_type), notch = TRUE) +
  labs(title = "Distribution of Sale Price for Residential Types",
       x = "Residential Type",
       y = "Sale Price") +
  theme_light()
```

__Counts of "Sale Price" facet wrapped by ""Property Type"__
```{r}
connecticut %>%
  ggplot(aes(x=sale_price, fill = property_type)) +
    geom_histogram(binwidth = 200000) +
    labs(title = "Distribution of Sale Price: Residential Type",
         ylab = "Count",
         xlab = "Sale Price") +
    theme_bw() +
    facet_wrap(vars(property_type), 
               scales = "free_y", 
                        "free_x")
```
* These graphics display a distribution that is still skewed to the right but
  that is more normally distributed than it was prior to being trimmed.

__Counts of "Sale Price" facet wrapped by ""Property Type"__
```{r}
connecticut %>%
  ggplot(aes(x=sale_price)) +
    geom_histogram(binwidth = 50000, fill = "steelblue") +
    labs(title = "Distribution of Sale Price",
         y = "Count",
         x = "Sale Price") +
    theme_bw() 
```



